---
title: "Unit 06: Normal Distribution & Central Limit Theorem"
subtitle: "Instructor: Dr. Md Mutasim Billah"

format:
  html:
    embed-resources: true
    toc: true

  pptx:
    slide-level: 2
    reference-doc: ../master-cwru.pptx
    
format-links: false
---


```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE, 
  warning = FALSE,
  message = FALSE,
  error = TRUE,
  fig.height = 2.75, 
  fig.width = 4.25,
  fig.env = 'figure',
  fig.pos = 'h',
  fig.align = 'center')
```


# Resources


## Handouts {-}

::: {.incremental}

- Download the [Unit 06 Handout: Normal Distribution & CTL](/handout/Unit-06-Normal-CTL.pdf) before class and bring it with you.

- In the class I have provided a copy of [Tables 1](/handout/Tables-1.pdf). If you have missed the class, meet me after the class or in mu OH to get a copy of the tables.

- I will continue to post additional handouts as we progress through the course. Please watch for **Canvas Announcements** when a new handout becomes available.

- It is very important that you **bring the appropriate handout to class regularly**, either as a printed copy or on your writing device.

:::




## Practice Problems {-}

::: {.incremental}

- **Practice Problems (PP)** are essential for strengthening your understanding, building confidence, and exploring the concepts covered in class.

- The **Unit 06 Practice Problems** will be posted soon

- Complete the relevant Practice Problems as we progress through the unit.

- Make sure you have completed and reviewed the Practice Problems **before attempting the corresponding HW assignment**.

:::


```{r normal-plot-setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE, warning = FALSE, message = FALSE,
  error = FALSE, fig.height = 2.75, fig.width = 4.25,
  fig.env = 'figure', fig.pos = 'h', fig.align = 'center'
)
normal_plot <- function(lower = -Inf, upper = Inf, mu = 0, sigma = 1,
                        shade = TRUE, marks = mu, labels = marks) {
  old <- par(mar = c(3, 2, 1, 1)); on.exit(par(old))
  xx <- seq(mu - 4*sigma, mu + 4*sigma, length.out = 1000)
  yy <- dnorm(xx, mu, sigma)
  plot(xx, yy, type = 'n', axes = FALSE, xlab = '', ylab = '',
       ylim = c(0, max(yy)*1.12))
  if (shade) {
    a <- max(min(xx), lower); b <- min(max(xx), upper)
    if (a < b) {
      x <- seq(a, b, length.out = 500)
      polygon(c(a, x, b), c(0, dnorm(x, mu, sigma), 0),
              col = '#8DE5ED', border = NA)
    }
  }
  lines(xx, yy, lwd = 2)
  axis(1, at = marks, labels = labels)
  abline(h = 0, col = 'gray40')
  mtext(if (mu == 0 && sigma == 1) 'z' else 'x', side = 1, line = 2)
}
```

<!-- Adapted from Philip Kendall's Introductory Statistics with R,
     Unit 05, Videos 05.05–05.07, supplied by the instructor.
     Repeated reveal slides have been combined. Graphs are regenerated in R.
     Table answers use four-decimal cumulative probabilities. -->

# Normal Distribution

## The Normal Distribution {-}

::: {.incremental}
- The **normal distribution**, often called the **bell curve**, is an important continuous probability distribution.
- Its support is $(-\infty,\infty)$.
- The curve is centered at its mean $\mu$; its spread depends on its standard deviation $\sigma>0$.
- We write $X\sim N(\mu,\sigma^2)$, using the **variance** as the second parameter. The source labels its curves using the mean and standard deviation explicitly.
:::

```{r normal-curve}
#| echo: false
normal_plot(shade = FALSE, marks = 0, labels = expression(mu))
```

## Normal Probability Density Function {-}

$$
f_X(x)=\frac{1}{\sqrt{2\pi}\,\sigma}
\exp\left[-\frac{(x-\mu)^2}{2\sigma^2}\right],
\qquad -\infty<x<\infty.
$$

The inflection points occur at $\mu-\sigma$ and $\mu+\sigma$.

```{r inflection-points}
#| echo: false
normal_plot(shade = FALSE, marks = c(-1,0,1),
            labels = expression(mu-sigma,mu,mu+sigma))
points(c(-1,1), dnorm(c(-1,1)), pch = 19, col = 'red')
```

## Comparing Normal Curves {-}

Changing $\mu$ shifts the center. Increasing $\sigma$ makes the curve wider and lowers its peak.

```{r compare-normal-curves}
#| echo: false
par(mar=c(3,3,1,1))
x <- seq(0,20,length.out=1000)
plot(x,dnorm(x,5,1),type='l',col='#2976B9',lwd=2,
     xlab='x',ylab='Density',ylim=c(0,.48))
lines(x,dnorm(x,15,1),col='#739B15',lwd=2)
lines(x,dnorm(x,5,2),col='#DE9500',lwd=2)
legend('top',c('Mean 5, SD 1','Mean 15, SD 1','Mean 5, SD 2'),
       col=c('#2976B9','#739B15','#DE9500'),lwd=2,bty='n',cex=.75)
```

## Example: Adult Heights {-}

Suppose adult men's heights are approximately normal with mean $\mu=70$ inches and standard deviation $\sigma=4$ inches.

How tall would someone be if their height were **10 standard deviations above the mean**?

$$
70+10(4)=110\text{ inches}=9\text{ feet }2\text{ inches}.
$$

This illustrates how extreme a value 10 standard deviations above the mean would be. For comparison, the source describes the tallest recorded man as approximately 8 feet 11 inches tall.

## The Empirical Rule {-}

For a normal distribution, approximately:

- **68%** of observations lie within $\mu\pm\sigma$;
- **95%** lie within $\mu\pm2\sigma$;
- **99.7%** lie within $\mu\pm3\sigma$.

```{r empirical-rule}
#| echo: false
normal_plot(-3,3,marks=-3:3,
 labels=expression(mu-3*sigma,mu-2*sigma,mu-sigma,mu,mu+sigma,mu+2*sigma,mu+3*sigma))
for (k in c(2,1)) {
 x <- seq(-k,k,length.out=300)
 polygon(c(-k,x,k),c(0,dnorm(x),0),border=NA,
         col=if(k==2) '#68DD9A' else '#FFF47A')
}
curve(dnorm(x),-4,4,add=TRUE,lwd=2)
text(0,c(.30,.12,.035),c('68%','95%','99.7%'))
```

## Empirical Rule: Heights and IQ Scores {-}

For the height model with $\mu=70$ and $\sigma=4$:

| Range | Interval (inches) | Approximate percentage |
|:--|:--|:--|
| $\mu\pm\sigma$ | 66–74 | 68% |
| $\mu\pm2\sigma$ | 62–78 | 95% |
| $\mu\pm3\sigma$ | 58–82 | 99.7% |

The source also illustrates an approximately normal IQ-score model with $\mu=100$ and $\sigma=15$. Its corresponding intervals are **85–115**, **70–130**, and **55–145**.

The source mentions *The Bell Curve* (Herrnstein and Murray, 1994) in this context. The normal model here is used only to illustrate the empirical rule.

# Standard Normal Distribution: Probabilities

## The Standard Normal Distribution {-}

There are infinitely many normal curves: $\mu$ may be any real number, and $\sigma$ may be any positive real number.

The **standard normal distribution** has $\mu=0$ and $\sigma=1$:

$$
Z\sim N(0,1),\qquad
f_Z(z)=\frac{1}{\sqrt{2\pi}}e^{-z^2/2},\quad -\infty<z<\infty.
$$

Its density has no elementary antiderivative. We use a **Z-table** or software to calculate areas.

## Reading the Z-Table {-}

::: {.incremental}
- The total area under the curve is **1**.
- The standard normal curve is symmetric about **0**.
- The table used here gives the area **to the left**: $\Phi(z)=P(Z\le z)$.
- Other tables may use different area conventions; always check the table heading.
- Use the row for the ones and tenths digits and the column for the hundredths digit.
- For a continuous random variable, $P(Z<z)=P(Z\le z)$.
:::

